---
title: "Datathon Final Submission"
author: "Suk Jin Chung, Maisara Mackeen"
format: 
  html:
    embed-resources: true
    self-contained: true
editor: visual
runtime: shiny
---

# Data Visualization

For this project, we chose the goverment prompt: **Analyzing the Seattle Resident's Customer Requests**

To better understand service demand patterns, we conducted a data visualization analysis of customer requests submitted by Seattle residents. By examining trends across different time periods and service categories, we identified key insights to support resource planning and operational improvements.

Our analysis is organized into three parts:

-   **Part 1:** Visualizing Customer Requests Over Time

-   **Part 2:** Identifying the Department with the Highest Number of Service Requests

-   **Part 3:** Analyzing Total Requests by Different Service Types

The data was sourced from [Seattle's Open Data Portal](https://data.seattle.gov/City-Administration/Customer-Service-Requests/5ngg-rpne/about_data).

## Part 1: Visualizing Customer Requests Over Time

We analyze customer service requests across different time periods to uncover patterns in activity:

-   **Month:** Identifies seasonal trends and highlights months with higher or lower request volumes.

-   **Quarter:** Groups requests into broader periods, revealing business or operational cycles.

-   **Day of Week:** Shows whether requests are more frequent on weekdays or weekends.

-   **Time (AM/PM):** Differentiates customer behavior between morning and afternoon/evening hours.

-   **Year:** Observes changes and growth in customer requests across different years.

By examining these trends, we can better understand when demand for services is highest and identify potential areas for resource planning or operational improvements.

```{r}
#| echo: false

library(ggplot2)
library(dplyr)
library(shiny)
library(readr)
library(scales)

#| echo: false
cleaned_df <- read_csv("Cleaned_Customer_Service.csv", show_col_types = FALSE)
```

```{r}
#| echo: false
#| input: true
selectInput(
  inputId = "time_unit",
  label = "Select Time Unit:",
  choices = c("Month", "Quarter", "DayOfWeek", "Time", "Year"),
  selected = "Month"
)

```

```{r}
#| echo: false
#| output: true
renderPlot({
  grouped_data <- cleaned_df %>%
    group_by(.data[[input$time_unit]]) %>%
    summarise(`Number of Requests` = n(), .groups = "drop") %>%
    arrange(.data[[input$time_unit]])
  
# If time unit is DayOfWeek, reorder manually
  if (input$time_unit == "DayOfWeek") {
    grouped_data[[input$time_unit]] <- factor(grouped_data[[input$time_unit]],
      levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
    )
  }

  ggplot(grouped_data, aes(x = .data[[input$time_unit]], y = `Number of Requests`)) +
    geom_bar(stat = "identity", fill = "darkblue") +
    scale_y_continuous(labels = comma) +
    labs(
      title = paste("Customer Requests by", input$time_unit),
      x = input$time_unit,
      y = "Number of Requests"
    )
})

```

```{r}
#| echo: false
#| output: true
renderUI({
  analysis_text <- switch(input$time_unit,
  "Month" = "Monthly trends reveal seasonal fluctuations in service demand. Identifying peak months helps departments anticipate and allocate resources proactively. Based on the bar chart above, <b>March</b> has the highest number of customer requests.",
  
  "Quarter" = "Grouping requests by quarter highlights business and operational cycles, supporting better alignment with fiscal planning and seasonal trends. Based on the bar chart above, the <b>first quarter</b> (January to March) has the highest number of customer requests.",
  
  "DayOfWeek" = "Viewing requests by day of the week shows whether demand is concentrated on weekdays or weekends, informing staffing and scheduling needs. Based on the bar chart above, <b>weekdays</b> receive the highest number of customer requests, with demand peaking on Monday and Tuesday and gradually declining as the week progresses.",
  
  "Time" = "Breaking down requests into AM and PM periods reveals daily behavioral patterns, helping optimize shift planning and response times. Based on the bar chart above, more customer requests are made during the <b>afternoon/evening</b> hours.",
  
  "Year" = "Analyzing requests across multiple years highlights long-term growth trends and major disruptions in service demand. Based on the bar chart above, customer requests have <b>steadily increased over the years</b>. Since we're just in April 2025 right now, we can anticipate that <b>2025</b> to experience even <b>higher</b> demand given this trend."
)

  
  HTML(paste0("<p><b>Analysis:</b> ", analysis_text, "</p>"))
})

```

## Part 2: Department with the Highest Number of Service Requests

This section identifies which city departments received the highest number of service requests. We summarized the total requests by department and visualized the results in a bar chart. The chart highlights which departments handle the greatest volume of requests, helping to reveal patterns in public service demand across the city.

```{r}
#| echo: false

#calculate number of requests for each department
department_counts <- cleaned_df %>%
  count(.data[["City Department"]]) %>%
  rename(`Number of Requests` = n) %>%
  arrange(desc(`Number of Requests`))

#plot graph for department_counts
ggplot(department_counts, aes(y = reorder(`City Department`, -`Number of Requests`), 
                              x = `Number of Requests`)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  #geom_text(aes(label = `Number of Requests`), vjust = -0.5, size = 3) + 
  scale_x_continuous(labels = comma) +
  labs(
    title = "Number of Service Requests by Department",
    x = "Number of Requests",
    y = "City Department"
  ) 

```

As seen in the bar chart above, SPD-Seattle Police Department has the highest number of service requests, which has a total of 256451 requests. This significant volume suggests that public safety concerns dominate service needs across the city.

To better understand the nature of these requests, we can further break down the types of issues reported to SPD.

```{r}
#| echo: false

# Filter for only SPD requests
spd_requests <- cleaned_df %>%
  filter(`City Department` == "SPD-Seattle Police Department")

# Count the types of requests
spd_request_types <- spd_requests %>%
  count(`Service Request Type`) %>%
  rename(`Number of Requests` = n) %>%
  arrange(desc(`Number of Requests`))

# View the top 3 types of service requests
top_spd_requests <- spd_request_types %>%
  slice_max(`Number of Requests`, n = 3)

# Plot the top 3 service request types for SPD
ggplot(top_spd_requests, aes(x = `Number of Requests`, 
                             y = reorder(`Service Request Type`, `Number of Requests`))) +
  geom_bar(stat = "identity", fill = "darkblue") +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Top 3 Service Request Types for SPD",
    x = "Number of Requests",
    y = "Service Request Type"
  ) +
  theme_minimal()

```

A closer breakdown of SPD requests shows that the two most common service types are:

-   **Parking Enforcement Requests** (ranked #1)

-   **Abandoned Vehicle/72-Hour Parking Ordinance Violations** (ranked #2)

These findings suggest that a large portion of public demand on SPD is related to **parking issues** rather than emergency or violent crime concerns.

Given this trend, **prioritization efforts** could include:

-   Strengthening **dedicated parking enforcement teams** to manage non-emergency parking complaints separately from police units.

-   **Public education campaigns** to clarify parking rules, aiming to reduce violations.

-   **Exploring technology solutions** (such as app-based parking violation reporting) to streamline response times.

Shifting non-emergency parking enforcement away from core police operations could improve SPD's capacity to focus on critical safety and crime prevention activities, while still addressing community concerns around parking and abandoned vehicles.

## Part 3: Total Request by Different Service Types

This section calculates the total requests that each service types obtained by the City of Seattle per quarter from the beginning of 2021 to 2025.

```{r}
#| echo: false

# Count instances of each service type requests by quarter
qtrly_request_counts <- count(cleaned_df, `Service Request Type`, Quarter, sort = TRUE)

# Get plot of the instances per quarter
ggplot(data = qtrly_request_counts, aes(x = `Service Request Type`, y = n, fill = Quarter)) + 
  geom_bar(stat = "identity") +
  labs(
    title = "Quarterly Count of Service Request Types",
    x = "Service Request Types", 
    y = "Count of Service Request Types", color = "Quarter"
    ) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Since there are a lot of service request types, the types are not that legible. The bar chart showing the counts of each type cannot be determined clearly – only the types with massive amounts of requests could be seen. Trends for each type could not be determined for certain due to the large number of data.

Therefore, we will get the top 10 service request type counts before doing a trend analysis for better graph legibility.

```{r}
#| echo: false
# GET TOP 10 FOR BETTER ANALYSIS

# Count instances of each service type, then sort and get the highest 10 service types
top_10_service_ct <- count(cleaned_df, `Service Request Type`, sort = TRUE) %>%
  slice_max(n, n = 10)

# Filter to the top 10 service types, then count instances of service type per quarter
top_10_qtrly_req_ct <- cleaned_df %>%
  filter(`Service Request Type` %in% top_10_service_ct$`Service Request Type`) %>%
  count(`Service Request Type`, Quarter)

# Plot in descending order
ggplot(data = top_10_qtrly_req_ct, aes(x = reorder(`Service Request Type`, -n), y = n, fill = Quarter)) +  
  geom_bar(stat = "identity") +
  labs(
    title = "Count of Top 10 Service Request Types per Quarter",
    x = "Service Request Types", 
    y = "Count of Service Request Types", 
    color = "Quarter") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

From the graph, **the top 10 service types**, in order, are:

1.  Abandoned Vehicle/72 hour Parking Ordinance

2.  Parking Enforcement

3.  Unauthorized Encampment

4.  Illegal Dumping / Needles

5.  Graffiti

6.  General Inquiry - Customer Service Bureau

7.  Pothole

8.  General Inquiry - Transportation

9.  Parks and Recreation Maintenance

10. Overgrown Vegetation

Over the quarters, there seems to be an increase in request types from quarter 1 to quarter 3, before decreasing in quarter 4. A lot of the service types like Unauthorized Encampment, Illegal Dumping/Needles, Graffiti and General Inquiry - Transportation have the highest amount of requests in quarter 3. For Pothole, there is a significant decline from quarter 1 to quarter 2 and quarter 2 to quarter 3, before it increases again in quarter 4. For Overgrown Vegetation, quarter 1 and 4 have noticeable lower number of requests compared to quarter 2 and 3. For the others, there are small decrease and increase of requests in between the quarters. Overall, the service type with the biggest amount of requests is Abandoned Vehicle/72-hr Parking Ordinance for every quarter.
