---
title: "Datathon Final Submission"
author: "Suk Jin Chung, Maisara Mackeen"
format: 
  html:
    embed-resources: true
editor: visual
runtime: shiny
---

# Data Visualization

## Part 1: Visualizing Customer Requests Over Time

```{r}
#| echo: false
library(ggplot2)
library(dplyr)
library(shiny)
library(readr)
library(scales)

cleaned_df <- read_csv("Cleaned_Customer_Service.csv", show_col_types = FALSE)
```

```{}
```

```{}
```

## Part 2: Department with the Highest Number of Service Requests

```{}
```

## Part 3: Total Request by Different Service Types

This section calculates the total requests that each service types obtained by the City of Seattle per quarter from the beginning of 2021 to 2025.

```{r}
#| echo: false
#| output: true

# Count instances of each service type requests by quarter
qtrly_request_counts <- dplyr::count(cleaned_df, `Service Request Type`, Quarter, sort = TRUE)

# Get plot of the instances per quarter
ggplot(data = qtrly_request_counts, aes(x = `Service Request Type`, y = n, fill = Quarter)) + 
  geom_bar(stat = "identity") +
  labs(
    title = "Quarterly Count of Service Request Types",
    x = "Service Request Types", 
    y = "Count of Service Request Types", color = "Quarter"
    ) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Since there are a lot of service request types, the types are not that legible. The bar chart showing the counts of each type cannot be determined clearly – only the types with massive amounts of requests could be seen. Trends for each type could not be determined for certain due to the large number of data.

Therefore, we will get the top 10 service request type counts before doing a trend analysis for better graph legibility.

```{r}
#| echo: false
# GET TOP 10 FOR BETTER ANALYSIS

# Count instances of each service type, then sort and get the highest 10 service types
top_10_service_ct <- dplyr::count(cleaned_df, `Service Request Type`, sort = TRUE) %>%
  dplyr::slice_max(n, n = 10)

# Filter to the top 10 service types, then count instances of service type per quarter
top_10_qtrly_req_ct <- cleaned_df %>%
  dplyr::filter(`Service Request Type` %in% top_10_service_ct$`Service Request Type`) %>%
  dplyr::count(`Service Request Type`, Quarter)

# Plot in descending order
ggplot(data = top_10_qtrly_req_ct, aes(x = reorder(`Service Request Type`, -n), y = n, fill = Quarter)) +  
  geom_bar(stat = "identity") +
  labs(
    title = "Count of Top 10 Service Request Types per Quarter",
    x = "Service Request Types", 
    y = "Count of Service Request Types", 
    color = "Quarter") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

From the graph, the top 10 service types, in order, are:

1.  Abandoned Vehicle/72 hour Parking Ordinance

2.  Parking Enforcement

3.  Unauthorized Encampment

4.  Illegal Dumping / Needles

5.  Graffiti

6.  General Inquiry - Customer Service Bureau

7.  Pothole

8.  General Inquiry - Transportation

9.  Parks and Recreation Maintenance

10. Overgrown Vegetation

Over the quarters, there seems to be an increase in request types from quarter 1 to quarter 3, before decreasing in quarter 4. A lot of the service types like Unauthorized Encampment, Illegal Dumping/Needles, Graffiti and General Inquiry - Transportation have the highest amount of requests in quarter 3. For Pothole, there is a significant decline from quarter 1 to quarter 2 and quarter 2 to quarter 3, before it increases again in quarter 4. For Overgrown Vegetation, quarter 1 and 4 have noticeable lower number of requests compared to quarter 2 and 3. For the others, there are small decrease and increase of requests in between the quarters. Overall, the service type with the biggest amount of requests is Abandoned Vehicle/72-hr Parking Ordinance for every quarter.
